---
title: "Br 8325 basic analysis"
author: "Sowmya"
date: "8/26/2021"
output: html_document
---



```{r setup, echo=FALSE, include = FALSE, message=FALSE}
library("here")
library("sessioninfo")
library("pryr")

## reading the data
library("SpatialExperiment")

## vis
library("spatialLIBD")

## analysis ## requires uwot for UMAP
library("scater")
library("BiocParallel")
library('ggspavis')

```


### build spe
```{r build spe, echo=FALSE}

dir <- 'D:/DLPFC images/8325 mid/OUTS'

samples <- file.path(dir)
sample_ids = 'Br8325_mid_manual_allignment_all'

spe_8325_mid <- read10xVisium(samples, sample_ids,
  type = "sparse",
  data = "raw",
  images = c("lowres", "hires", "fullres", "detected", "aligned"),
  load = TRUE
)
```

### add counts data
```{r add counts data, echo = FALSE}

counts = read.csv('D:/DLPFC images/8325 mid/OUTS/spatial/tissue_spot_counts.csv')
colData(spe_8325_mid)$cell_counts = counts$cell_counts
colData(spe_8325_mid)$image_row  = counts$imagerow
colData(spe_8325_mid)$image_col  = counts$imagecol
  

```

### plot spots
```{r plot spots}

plotSpots(spe_8325_mid,
  x_coord = "pxl_row_in_fullres",
  y_coord = "pxl_col_in_fullres")

```

## Quality Control

Low-quality spots can be identified according to several characteristics, including:

1. library size (i.e. total UMI counts per spot)

2. number of expressed features (i.e. number of genes with non-zero UMI counts per spot)

3. proportion of reads mapping to mitochondrial genes (a high proportion indicates cell damage)

4. number of cells per spot (unusually high values can indicate problems)


### keep only spots over tissue
```{r keep only spots over tissue, echo=FALSE}

spe_8325_mid <- spe_8325_mid[, spatialData(spe_8325_mid)$in_tissue == 1]
dim(spe_8325_mid)


```

### identify mitochondrial genes
```{r identify mitochondrial genes, echo=FALSE}

is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe_8325_mid)$symbol)
table(is_mito)


```

###  list MT genes
``` {r list MT genes, echo = FALSE}
rowData(spe_8325_mid)$symbol[is_mito]
```
### Calculate per-spot QC metrics and store in colData
```{r Calculate per-spot QC metrics and store in colData, echo = FALSE}

spe_8325_mid <- addPerCellQC(spe_8325_mid, subsets = list(mito = is_mito))
```

### library size histogram
```{r library size histogram, echo = FALSE}

hist(colData(spe_8325_mid)$sum, breaks = 40)

```
Spike at low library size?  

### plot cell counts
``` {r plot cell counts, echo = FALSE}

plotQC(spe_8325_mid, type = "scatter", 
       metric_x = "cell_counts", metric_y = "sum", 
       threshold_y = 500)

```

### spatial pattern of discarded spots for high threshold
``` {r spatial pattern of discarded spots for high threshold}
qc_lib_size_2000 <- colData(spe_8325_mid)$sum < 2000
colData(spe_8325_mid)$qc_lib_size_2000 <- qc_lib_size_2000

plotQC(spe_8325_mid, type = "spots", 
       x_coord = "pxl_row_in_fullres",
       y_coord = "pxl_col_in_fullres",
       discard = "qc_lib_size_2000")
```


``

###expressed genes
```{r expressed genes }
#histogram of numbers of expressed genes
hist(colData(spe_8325_mid)$detected, breaks = 20)
```

### num expressed genes vs. num cells per spot
```{r num expressed genes vs. num cells per spot}

plotQC(spe_8325_mid, type = "scatter", 
       metric_x = "cell_counts", metric_y = "detected", 
       threshold_y = 250)

```

### threshold for expressed genes
```{r threshold for expressed genes}

qc_detected <- colData(spe_8325_mid)$detected < 250

colData(spe_8325_mid)$qc_detected <- qc_detected

plotQC(spe_8325_mid, type = "spots", 
       x_coord = "pxl_row_in_fullres",
       y_coord = "pxl_col_in_fullres",
       discard = "qc_detected")

```

### check spatial pattern for high threshold
```{r check spatial pattern for high threshold}

qc_detected_1000 <- colData(spe_8325_mid)$detected < 1000
colData(spe_8325_mid)$qc_detected_1000 <- qc_detected_1000
plotQC(spe_8325_mid, type = "spots", 
       x_coord = "pxl_row_in_fullres",
       y_coord = "pxl_col_in_fullres",
       discard = "qc_detected_1000")


```

## proportion of mitochondrial reads 

``` {r mito histogram}

hist(colData(spe_8325_mid)$subsets_mito_percent, breaks = 20)

```

### mito proportion vs. num cells per spot
``` {r mito proportion vs. num cells per spot}

plotQC(spe_8325_mid, type = "scatter",
       x_coord = "pxl_row_in_fullres",
       y_coord = "pxl_col_in_fullres",
       metric_x = "cell_counts", metric_y = "subsets_mito_percent", 
       threshold_y = 30)

```

### mito thresholding @ 30
```{r mito thresholding @ 30}

qc_mito <- colData(spe_8325_mid)$subsets_mito_percent > 30
table(qc_mito)
colData(spe_8325_mid)$qc_mito <- qc_mito

plotQC(spe_8325_mid, type = "spots", 
       x_coord = "pxl_row_in_fullres",
       y_coord = "pxl_col_in_fullres",
       discard = "qc_mito")

```

### cell counts histogram
```{r cell counts histogram}

hist(colData(spe_8325_mid)$cell_counts, breaks = 20)

```


### r distribution of cells per spot
```{r distribution of cells per spot}
tbl_cells_per_spot <- table(colData(spe_8325_mid)$cell_count)
```

### num expressed genes vs. thresholded num cells
```{r num expressed genes vs. thresholded num cells}
plotQC(spe_8325_mid, type = "scatter", 
       metric_x = "cell_counts", metric_y = "detected", 
       threshold_x = 12)


```

From 9.5.4: 

spots with very high cell counts also have low numbers of expressed genes. This indicates that the experiments have failed for these spots, and they should be removed.

Is 12 still a good threshold? 

### selected QC threshold for number of cells per spot
```{r selected QC threshold for number of cells per spot }

qc_cell_count <- colData(spe_8325_mid)$cell_counts > 12
table(qc_cell_count)

colData(spe_8325_mid)$qc_cell_count <- qc_cell_count

# check spatial pattern of discarded spots
plotQC(spe_8325_mid, type = "spots", 
       x_coord = "pxl_row_in_fullres",
       y_coord = "pxl_col_in_fullres",
       discard = "qc_cell_count")
```


Number of cells correlated with wrinkles. But is the other part of the tissue damaged? High mito count? 

Should I be discarding anything at all? 

## Normalization Chapter 10 

Cluster similar cells based on their expression profiles, using either log-expression values or ranks.
```{r quick clustering }
library(scran)
set.seed(123)
qclus <- quickCluster(spe_8325_mid)
table(qclus)

```

### store size factors in objects
```{r store size factors in objects}

# calculate size factors and store in object
spe_8325_mid <- computeSumFactors(spe_8325_mid, cluster = qclus)

summary(sizeFactors(spe_8325_mid))

```

### size factor hist 
```{r size factor hist }
hist(sizeFactors(spe_8325_mid), breaks = 20)
```


### calculate logcounts (log-transformed normalized counts)
```{r calculate logcounts (log-transformed normalized counts)}

spe_8325_mid<- logNormCounts(spe_8325_mid)

# check
assayNames(spe_8325_mid)
```

## Feature selection and Dimensionality reduction 

```{r feature selection}

library(scran)
# remove mitochondrial genes
spe_8325_mid <- spe_8325_mid [!is_mito, ]
# fit mean-variance relationship
dec <- modelGeneVar(spe_8325_mid)
# select top HVGs
top_hvgs <- getTopHVGs(dec, prop = 0.1)

# visualize mean-variance relationship
fit <- metadata(dec)
plot(fit$mean, fit$var, 
     xlab = "mean of log-expression", ylab = "variance of log-expression",
     main = '8325 mid')
curve(fit$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)

```
Apply principal component analysis (PCA) to the set of top highly variable genes (HVGs) to reduce the dimensionality of the dataset, and retain the top 50 principal components (PCs) for further downstream analyses.

### compute PCA
```{r compute PCA}

set.seed(123)
spe_8325_mid <- runPCA(spe_8325_mid, subset_row = top_hvgs)

reducedDimNames(spe_8325_mid)
dim(reducedDim(spe_8325_mid, "PCA"))
```

### UMAP
```{r UMAP}
library(uwot)
set.seed(123)
spe_8325_mid <- runUMAP(spe_8325_mid, dimred = "PCA")

reducedDimNames(spe_8325_mid)
dim(reducedDim(spe_8325_mid, "UMAP"))
colnames(reducedDim(spe_8325_mid, "UMAP")) <- paste0("UMAP", 1:2)

```

```{r plot top 2 PCA dimensions}
plotDimRed(spe_8325_mid, type = "PCA")
```

```{r plot top 2 UMAP dimensions}
plotDimRed(spe_8325_mid, type = "UMAP")
```

##Clustering 
```{r graph based clustering}
# graph-based clustering
set.seed(123)
k <- 10
g <- buildSNNGraph(spe_8325_mid, k = k, use.dimred = "PCA")
g_walk <- igraph::cluster_walktrap(g)
clus <- g_walk$membership
table(clus)

# store cluster labels in column 'label' in colData
colLabels(spe_8325_mid) <- factor(clus)

plotSpots(spe_8325_mid, annotate = "label",
          x_coord = "pxl_row_in_fullres",
          y_coord = "pxl_col_in_fullres",
          palette = "libd_layer_colors")


```
```{r plot labeled PCA}
# plot clusters in PCA reduced dimensions
plotDimRed(spe_8325_mid, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors")
```
```{r plot labeled UMAP}
# plot clusters in PCA reduced dimensions
plotDimRed(spe_8325_mid, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors")
```

## Identifying Marker Genes 

```{r gene per cluster}
# set gene names as row names for easier plotting
rownames(spe_8325_mid) <- rowData(spe_8325_mid)$symbols

# test for marker genes
markers <- findMarkers(spe_8325_mid, test = "binom", direction = "up")

# returns a list with one DataFrame per cluster
markers

```


```{r log-fold changes for clusters}
# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1

library(pheatmap)
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)

pheatmap(logFCs, breaks = seq(-5, 5, length.out = 101))

```
``` {r plot mito vs sum, echo = FALSE}
plotQC(spe_8325_mid, type = "scatter", 
       metric_x = "detected", metric_y = "subsets_mito_percent")
```


